Residualized Change Score Example

### 2023-04-19 1a_Difference Scores paired samples ttest_HowTo.qmd
rstudio test commit 3
rstudio test commit 4

---
format: 
  #html:
  #  toc: true
  #  code-fold: true
  docx:
    toc: true
prefer-html: true
---

@@

## Preliminary Setup

::: panel-tabset
### Install Packages {.tabset .tabset-fade .tabset-pills}

::: blue
> **This code installs the r packages necessary for this example, if
> they are not already installed**

```{r pckg-install}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE

#rm(list = ls())

if (!("lme4" %in% installed.packages())) install.packages("lme4")
if (!("lmerTest" %in% installed.packages())) install.packages("lmerTest")
if (!("tidyverse" %in% installed.packages())) install.packages("tidyverse")
if (!("arrow" %in% installed.packages())) install.packages("arrow")
if (!("arsenal" %in% installed.packages())) install.packages("arsenal")
if (!("kableExtra" %in% installed.packages())) install.packages("kableExtra")
if (!("ggpubr" %in% installed.packages())) install.packages("ggpubr")
if (!("rstatix" %in% installed.packages())) install.packages("rstatix")

```
:::

### Load Packages

::: blue
**This code loads the r libraries necessary for this example**

```{r lib-load}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| class.source: fold-hide

#rm(list = ls())

library(lme4)
library(lmerTest)
library(tidyverse)
library(arrow)
library(arsenal)
library(kableExtra)
library(rstatix)
library(ggpubr)
```
:::

### Config Options

::: blue
**This code configures knitr code chunk options**

```{r config}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE

knitr::opts_chunk$set(echo = T, message=F, warning=F, error=F, 
                                  comment=NA, cache=T, code_folding=T,
                                  R.options=list(width=220), fig.align='center', 
                                  out.width='75%', fig.asp=.75)
```
:::
:::

## Descriptives Overview

::: panel-tabset
### Read and View Data {.tabset .tabset-fade .tabset-pills}

::: blue
**This code reads in and shows the data to be used in the current
example**

```{r read-data}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE

## Read data
df_long<- read_csv("/Users/shawes/Desktop/data/df_long.csv")
df_wide <- read_csv("/Users/shawes/Desktop/data/df_wide.csv")

# set types
df_long$ids <- as.factor(df_long$ids)
df_long$event <- as.factor(df_long$event)
df_long$site_id <- as.factor(df_long$site_id)
df_long$fam_id <- as.factor(df_long$fam_id)
df_long$sex <- as.factor(df_long$sex)
df_long$age <- as.numeric(df_long$age)
df_long$mature_vg <- as.numeric(df_long$mature_vg)
df_long$vg_wkdy <- as.numeric(df_long$vg_wkdy)
df_long$vg_wknd <- as.numeric(df_long$vg_wknd)
df_long$cbcl_extern <- as.numeric(df_long$cbcl_extern)
df_long$cbcl_cd <- as.numeric(df_long$cbcl_cd)
df_long$ksad_cdco1 <- as.factor(df_long$ksad_cdco1)
df_long$vg_total <- as.numeric(df_long$vg_total)
```
:::

### Continuous Outcomes

::: blue
**This code creates a table of descriptive information for continuous
outcomes**

```{r contvar-table}
#| messages: FALSE
#| echo: TRUE
#| warning: FALSE
#| results: "asis"

## Create a descriptives table of study variables by measurement occasion 
tab_descriptives_1 <- tableby.control(test=FALSE, total=FALSE, numeric.test="kwt", cat.test="chisq",
numeric.stats=c("N", "meansd", "median", "range"), #"Nmiss2"
cat.stats=c("countpct"), #"Nmiss2"
stats.labels=list(N='Count', meansd="Mean (SD)", median ='Median', 
                  range='Min - Max'))  #, Nmiss2 ='Missing'
my_cont_labels <- list(age = "Age", vg_total = "Weekly # of Video Gaming Hrs",
            cbcl_extern = "CBCL Externalizing Scale"
)

tab_descriptives_1 <- tableby(event ~ age + vg_total + 
                                      cbcl_extern,
                                      data=df_long, control=tab_descriptives_1)

# Push table object through kable and kable_styling
tab_descriptives_1 %>%
  summary(text=TRUE, digits.pct=1, digits=1) %>%
  kable(caption = "Continuous Outcomes") %>%
  kable_styling(bootstrap_options = "striped", 
                full_width = FALSE, html_font = "Cambria",
                font_size = 15, position = "center", fixed_thead = T) %>%
  row_spec(2:3,  bold = F, extra_css = 'vertical-align: middle !important;') %>% 
  column_spec(1, width = "20em", background = "light grey", bold = T, border_right = T) %>%
  column_spec(2, width = "20em", border_right = T) %>%
  column_spec(3, width = "20em", border_right = T) %>%
  footnote(general = "Here is a general comments of the table. ") %>%
  scroll_box(width = "75%", height = "500px")
```
:::

### Categorical Outcomes

::: blue
**This code creates a descriptives table for categorical outcomes**

```{r catvar-table}
#| class.source: 'fold-hide'
#| messages: FALSE
#| warning: FALSE
#| results: "asis"

## Create a descriptives table of study variables by measurement occasion 
tab_descriptives_2  <- tableby.control(test=FALSE, total=FALSE,
numeric.test="kwt", cat.test="chisq",
numeric.stats=c("N", "meansd", "median", "range"
), # "Nmiss2"
cat.stats=c("countpct"), # "Nmiss2"
stats.labels=list(N='Count', meansd="Mean (SD)", median
='Median', range='Min - Max'
)) # , Nmiss2 ='Missing'

my_cat_labels  <- list(
  event = "Year",
  sex = "Sex",
  mature_vg = "Mature Video Games"
  )

tab_descriptives_2 <- tableby(event ~ sex + mature_vg, data=df_long, control=tab_descriptives_2)

# Push table object through kable and kable_styling
tab_descriptives_2 %>%
            summary(text=TRUE, digits.pct=1, digits=1) %>%
            kable(caption = "Categorical Outcomes") %>%
            kable_styling(bootstrap_options = "striped", full_width = FALSE, font_size =
                            15, position = "center", fixed_thead = T) %>%
  row_spec(2:3,  bold = F, extra_css = 'vertical-align: middle !important;') %>%
  column_spec(1, width = "20em", background = "light grey", bold = T, border_right = T) %>%
  column_spec(2, width = "20em", border_right = T) %>%
  #column_spec(3, width = "20em", border_right = T) %>%
  footnote(general = "Here is a general comments of the table. ") %>%
  scroll_box(width = "75%", height = "500px")

```
:::
:::

## Results

<div class = "blue">

::: panel-tabset
### Build Model {.tabset .tabset-fade .tabset-pills}

::: blue
The code snippet below tells R to compute a difference score by
subtracting each participant's Externalizing score at T2 from their
Externalizing score at T1. Next, a one-sample t-test is computed to
examine whether the average difference score is different than zero
(indicating change).

**STEP 1: Compute Difference Score**

```{r build-model-1}
#| class.source: 'fold-hide'
#| message: FALSE
#| warning: FALSE
#| echo: TRUE

# Compute difference score based on CBCL Externalizing subscale scores at baseline (t1) and 1-Year Follow-up (t2)
df_wide$diffscore <- df_wide$cbcl_extern_Year_1 - df_wide$cbcl_extern_Baseline
```

```{r model-checking}
# Compute statistical summaries for the difference score variable
summary(df_wide$diffscore)

```

This summary of the difference score variable indicates xxxxx.

```{r}
hist(df_long$vg_total)

## Summary statistics
summary<-df_long %>%
group_by(event) %>%
get_summary_stats(cbcl_extern, type = "mean_sd")
data.frame(summary)
```

This histogram indicates xxxxxx. The summary command shows xxxxxx.

```{r shapiro-wilks}
### Shapiro-Wilk test and normality (Q-Q) plot (visualization of correlation between a given sample and the normal distribution)
shapiro.test(df_wide$cbcl_extern_Baseline[0:5000])
```

This Shapiro-Wilkes test shows xxxxxx.

```{r qqplot}
qqplot <- ggqqplot(df_wide$cbcl_extern_Baseline,
    ylab = "Externalizing Difference Score", xlab = FALSE,
    ggtheme = theme_minimal()
)

suppressWarnings(print(qqplot)) 
```

This qqplot shows xxxxxx.

**STEP 2: Conduct t-test on Difference Score**

```{r build-model-2}
#| class.source: 'fold-hide'
#| message: FALSE
#| warning: FALSE
#| echo: TRUE

# One-sample t-test to determine whether the average difference score is significantly different than 0.
result <- t.test(df_wide$cbcl_extern_Baseline, mu = 0, alternative = "two.sided")
# Print the results
result
```
:::

The output from our model provides: i. a t-statistic ii. degrees of
freedom under iii. p-value (\< .05 then we reject the null hypothesis)
iv. a mean estimate v. 95% confidence intervals

In our example p = .002 which is less than .05 then we reject the null
hypothesis and conclude that the mean difference score µ is different
from zero. So we can conclude that there𝐷 is a difference between cbcl
externalizing scores from t1 to t2. This output also provides the mean
for cbcl externalizing scores (4.45),as well as the correpsonding
confidence intervals

### Model Plots

::: blue
```{r scatterplot}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| code-summary: testing

# Scatterplot to visualize relationship between t1 & t2 data used to create difference score
scatterplot <- ggplot(df_wide, aes(x = cbcl_extern_Baseline, y = cbcl_extern_Year_1)) +
    geom_point(size = 3) +
    geom_smooth(method = lm, se = F) +
    xlab("CBCL Externalizing (x) Baseline") +
    ylab("CBCL Externalizing (x) 1-Year Follow-up")

suppressWarnings(print(scatterplot))
```

Examination of this scatterplot indicates xxxxx.

```{r boxplot}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| code-summary: testing

### boxplot
diffscore_boxplot <- ggboxplot(df_wide$cbcl_extern_Baseline,
    ylab = "Externalizing Difference Score", xlab = FALSE,
    ggtheme = theme_minimal()
)

suppressWarnings(print(diffscore_boxplot))
```

Examination of this boxplot indicates xxxxx.
:::

> This output shows xxxxxxxx. Briefly walk through each metric
:::

## Wrapping Up

<div class = "blue">

::: panel-tabset
### Write-up {.tabset .tabset-fade .tabset-pills}

::: blue
A paired samples t test was conducted to compare differences in
participants cbcl externalzing scores measured across two timepoints.
Findings indicated that there was a difference between cbcl externalzing
scores obtained at T1 compared to cbcl externalzing scores obtained at
T2, t(179) = 3.10, p = .002, Cohen's d = 0.23. Youth exhibited an
increase of cbcl externalzing scores at t2 ( M = 5.67, SD = 1.24)
compared to t1 ( M = 5.83, SD = 1.21).
:::
:::

>>>>

---
title: "DiffScoreExample"
---

@@

### 2023-04-19 2b_Residualized Change Score multiple regression_HowTo.qmd
rstudio test commit 3
rstudio test commit 4

---
format: 
  html:
    toc: true
    code-fold: true
  docx:
    toc: true
prefer-html: true
---

# Mulitple Regression {.unnumbered}

@@

## Preliminary Setup
::: {.panel-tabset}
### Install Packages {.tabset .tabset-fade .tabset-pills}
<div class = "blue">

> **This code installs the r packages necessary for this example, if they are not already installed**

```{r}
#| class.source: 'fold-hide'
#| message: FALSE
#| warning: FALSE
#| echo: FALSE

#rm(list = ls())

# Install necessary packages (if not already installed)
if (!("rstatix" %in% installed.packages())) install.packages("rstatix")
if (!("lme4" %in% installed.packages())) install.packages("lme4")
if (!("lmerTest" %in% installed.packages())) install.packages("lmerTest")
if (!("tidyverse" %in% installed.packages())) install.packages("tidyverse")
if (!("ggpubr" %in% installed.packages())) install.packages("ggpubr")
if (!("emmeans" %in% installed.packages())) install.packages("emmeans")
```
</div>

### Load Packages
<div class = "blue">

> **This code loads the r libraries necessary for this example**

```{r pckg-load}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| class.source: fold-hide

#rm(list = ls())

#Load packages
library(rstatix)
library(lme4)
library(lmerTest)
library(tidyverse)
library(ggpubr)
library(emmeans)
```

</div>

### Config Options
<div class = "blue">

**This code configures knitr code chunk options**

```{r config}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE

knitr::opts_chunk$set(echo = T, message=F, warning=F, error=F, 
                      comment=NA, cache=T, code_folding=T,
                      R.options=list(width=220), fig.align='center',
                      out.width='75%', fig.asp=.75)
```
</div>
:::

## Descriptives Overview
::: {.panel-tabset}
### Read and View Data {.tabset .tabset-fade .tabset-pills}
<div class = "blue">

**This code reads in and shows the data to be used in the current example**

```{r read-data}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE

## Read data
df_long<- read_csv("/Users/shawes/Desktop/data/df_long.csv")
df_wide <- read_csv("/Users/shawes/Desktop/data/df_wide.csv")

# set types (if necessary)
df_long$ids <- as.factor(df_long$ids)
df_long$event <- as.factor(df_long$event)
df_long$site_id <- as.factor(df_long$site_id)
df_long$fam_id <- as.factor(df_long$fam_id)
df_long$sex <- as.factor(df_long$sex)
df_long$age <- as.numeric(df_long$age)
df_long$mature_vg <- as.numeric(df_long$mature_vg)
df_long$vg_wkdy <- as.numeric(df_long$vg_wkdy)
df_long$vg_wknd <- as.numeric(df_long$vg_wknd)
df_long$cbcl_extern <- as.numeric(df_long$cbcl_extern)
df_long$cbcl_cd <- as.numeric(df_long$cbcl_cd)
df_long$ksad_cdco1 <- as.factor(df_long$ksad_cdco1)
df_long$vg_total <- as.numeric(df_long$vg_total)
```

</div>

:::

### Continuous Outcomes

::: blue
**This code creates a table of descriptive information for continuous
outcomes**

```{r contvar-table}
#| messages: FALSE
#| echo: TRUE
#| warning: FALSE
#| results: "asis"

## Create a descriptives table of study variables by measurement occasion 
tab_descriptives_1 <- tableby.control(test=FALSE, total=FALSE, numeric.test="kwt", cat.test="chisq",
numeric.stats=c("N", "meansd", "median", "range"), #"Nmiss2"
cat.stats=c("countpct"), #"Nmiss2"
stats.labels=list(N='Count', meansd="Mean (SD)", median ='Median', 
                  range='Min - Max'))  #, Nmiss2 ='Missing'
my_cont_labels <- list(age = "Age", vg_total = "Weekly # of Video Gaming Hrs",
            cbcl_extern = "CBCL Externalizing Scale"
)

tab_descriptives_1 <- tableby(event ~ age + vg_total + 
                                      cbcl_extern,
                                      data=df_long, control=tab_descriptives_1)

# Push table object through kable and kable_styling
tab_descriptives_1 %>%
  summary(text=TRUE, digits.pct=1, digits=1) %>%
  kable(caption = "Continuous Outcomes") %>%
  kable_styling(bootstrap_options = "striped", 
                full_width = FALSE, html_font = "Cambria",
                font_size = 15, position = "center", fixed_thead = T) %>%
  row_spec(2:3,  bold = F, extra_css = 'vertical-align: middle !important;') %>% 
  column_spec(1, width = "20em", background = "light grey", bold = T, border_right = T) %>%
  column_spec(2, width = "20em", border_right = T) %>%
  column_spec(3, width = "20em", border_right = T) %>%
  footnote(general = "Here is a general comments of the table. ") %>%
  scroll_box(width = "75%", height = "500px")
```  
:::

### Categorical Outcomes

::: blue
**This code creates a descriptives table for categorical outcomes**

```{r catvar-table}
#| class.source: 'fold-hide'
#| messages: FALSE
#| warning: FALSE
#| results: "asis"

## Create a descriptives table of study variables by measurement occasion 
tab_descriptives_2  <- tableby.control(test=FALSE, total=FALSE,
numeric.test="kwt", cat.test="chisq",
numeric.stats=c("N", "meansd", "median", "range"
), # "Nmiss2"
cat.stats=c("countpct"), # "Nmiss2"
stats.labels=list(N='Count', meansd="Mean (SD)", median
='Median', range='Min - Max'
)) # , Nmiss2 ='Missing'

my_cat_labels  <- list(
  event = "Year",
  sex = "Sex",
  mature_vg = "Mature Video Games"
  )

tab_descriptives_2 <- tableby(event ~ sex + mature_vg, data=df_long, control=tab_descriptives_2)

# Push table object through kable and kable_styling
tab_descriptives_2 %>%
            summary(text=TRUE, digits.pct=1, digits=1) %>%
            kable(caption = "Categorical Outcomes") %>%
            kable_styling(bootstrap_options = "striped", full_width = FALSE, font_size =
                            15, position = "center", fixed_thead = T) %>%
  row_spec(2:3,  bold = F, extra_css = 'vertical-align: middle !important;') %>%
  column_spec(1, width = "20em", background = "light grey", bold = T, border_right = T) %>%
  column_spec(2, width = "20em", border_right = T) %>%
  #column_spec(3, width = "20em", border_right = T) %>%
  footnote(general = "Here is a general comments of the table. ") %>%
  scroll_box(width = "75%", height = "500px")

```
:::
:::

## Results

<div class = "blue">

::: panel-tabset
### Build Model {.tabset .tabset-fade .tabset-pills}

::: blue

The code snippet below tells R to compute xxxxx by xxxxxxx. Next, xxxxx is conducted to examine whether xxxxxx does xxxxxxx (indicating xxxxx). The xxxx function (xxx) is provided by the r 'xxx' package.

**STEP 1: Compute xxxxxxxx**
```{r build-model-1}
#| class.source: 'fold-hide'
#| message: FALSE
#| warning: FALSE
#| echo: TRUE

#A multiple regression approach can be used to examine effects of a continuous covariate  when controlling for autoregressive effects (t1->t2)

#Multiple regression model
result = lm(cbcl_extern_Year_1 ~ cbcl_extern_Baseline + vg_total_Baseline, data = df_wide)
str(df_wide)
# fit ML model including mean structure to make comparable with FIML fit below
# (means are always included with FIML model fits)
# sem.mod.fit <- sem(sem.mod, data=tpb.df, meanstructure=TRUE)

# fit again including missing data also
#sem.mod.fit.fiml <- sem(sem.mod, data=tpb.df, missing="ML")

```

```{r model-checking}
#Diagnostic Plots

summary(result)

# residuals vs. fitted values plot.
resid_plot<- plot(result, add.smooth = FALSE, which = 1)

# normal probability plot
norm_plot<- plot(result, which = 2)

# scale location plot
scale_plot<- plot(result, add.smooth = FALSE, which = 3)

# boxplot
#boxplot <- ggplot(df_wide, aes(x = Externalizing_T1, y = Externalizing_T0, colour = Sex)) + geom_point()

#suppressWarnings(print(resid_plot))
#suppressWarnings(print(norm_plot))
#suppressWarnings(print(scale_plot))
#suppressWarnings(print(boxplot))

```

This summary of xxxx indicates xxxxx.

The output from our model provides: i. xxxx; ii. xxxxx; iii. xxxx. In our example xxxxx indicates xxxxx. An examination of xxxxx is xxxx, indicating xxxxxx. Then we conclude that xxxxxx.

```{r}
# Boxplots

df_wide %>%
            group_by(Sex) %>%  
            summarise(mean_grade = mean(Externalizing_T0),
            sd_grade = sd(Externalizing_T0),
            mean_exam = mean(Externalizing_T1),
            sd_exam = sd(Externalizing_T1))

# Boxplot of CBCL Externalizing scores at baseline by sex
boxplot_cbclExt_T1 <- boxplot(Externalizing_T0 ~ Sex,data = df_wide,
main = "CBCL Ext Baseline Score by sex",
xlab = "Sex",ylab = "Baseline CBCL Externalizing Score",
col = "red",border = "black")

# Boxplot of CBCL Externalizing scores at 1-Year follow-up by sex
boxplot_cbclExt_T2 <- boxplot(Externalizing_T1 ~ Sex, data = df_wide,
main = "CBCL Ext 1-Year Follow-up Score by sex",
xlab = "Sex",ylab = "1-Year Follow-up CBCL Externalizing Score",
col = "red",border = "black")

suppressWarnings(print(boxplot_cbclExt_T1))
suppressWarnings(print(boxplot_cbclExt_T2))
```

These boxplots indicates xxxxxx.

## Wrapping Up
<div class = "blue">

::: panel-tabset
### Write-up {.tabset .tabset-fade .tabset-pills}

::: blue

A xxxxxx analysis was conducted to examine xxxxx. Findings showed xxxxxx

:::
:::

>>>>

---
title: "Residualized Change Score Example"
---

# Mulitple Regression
testing testing testing

@@
